---
title: "Comparing GPT-4.1 Models for Screening"
format:
  html:
    code-fold: false
    code-tools: true
    code-summary: "Show the code"
    toc: true
execute: 
  echo: fenced
  warning: false
  message: false
knitr:
  opts_chunk: 
    fig.pos: "H"
    fig.retina: 2
    cache: FALSE
    R.options:
      knitr.graphics.auto_pdf: true
      width: 100
      knitr.kable.NA: "-"
      dplyr.summarise.inform: FALSE
      scipen: 10
      pillar.sigfig: 4 
editor_options: 
  chunk_output_type: console
---

## Introduction

The `AIscreenR` package supports several OpenAI models for screening, including `gpt-4.1`, `gpt-4.1-mini`, and `gpt-4.1-nano`.
In this article, we compare the performance of these models on a real-world dataset to evaluate their effectiveness in systematic literature review screening tasks. 
We will analyze their performance with a single repetition (`reps = 1`). For the faster and more economical `gpt-4.1-mini` and `gpt-4.1-nano` models, we will also investigate the impact of increasing the screening to ten repetitions (`reps = 10`).

## Setup

First, we load the necessary packages for our analysis.

```{r}
#| echo: true
#| eval: false
library(AIscreenR)
library(dplyr)
library(knitr)
```

## Methodology: Running the Comparison

To compare the models, we performed a screening task on the `friends_dat` dataset. The goal was to identify studies about the FRIENDS preventive programme.

The following prompt was used for all models. It is designed to be very specific about the required output format to ensure consistent results.

```{r}
#| echo: true
#| eval: false

# This is the prompt used for the experiment.
prompt <- "You are a systematic literature review screening agent. You are equipped with a screening function tool that you MUST use.

MANDATORY BEHAVIOR:
- Call the provided screening function tool immediately after reading each study
- Do not provide any response other than the function call
- The function call is your complete and only response

SCREENING CONTEXT:
Topic: FRIENDS preventive programme's effect on reducing anxiety symptoms in children/adolescents
Criteria 1: Study about FRIENDS preventive programme. The FRIENDS programme is a 10-session manualised cognitive behavioural therapy (CBT) programme which can be used as both prevention and treatment of child and youth anxiety. 
Criteria 2: Study is estimating an effect between a treatment and control/comparison group

WORKFLOW:
1. Read title and abstract
2. Make function call with your assessments
3. End response immediately after function call

CRITICAL: Your entire response must be the function tool call. No additional text before, during, or after the function call."
```

For each model (`gpt-4.1`, `gpt-4.1-mini`, `gpt-4.1-nano`), we ran the screening using `tabscreen_gpt()`. We ran the process twice: once with a single repetition (`reps = 1`) and once with ten repetitions (`reps = 10`) (only for the models `gpt-4.1-mini`and `gpt-4.1-nano`).

The code to run the screening for a single model looked like this:

```{r}
#| echo: true
#| eval: false

# This code was run for each model and repetition setting.
# It is not evaluated here to save time, as we will load the results directly.
plan(multisession)

results <- 
  tabscreen_gpt(
    data = friends_dat, 
    prompt = prompt, 
    model = "gpt-4.1", # This was changed for each model
    reps = 1 # This was changed to 10 for the second run
  )

plan(sequential)

# The results were saved to an .rds file for later analysis.
saveRDS(results, "result_obj_gpt-4.1_rep1.rds")
```

## Results

We now load the pre-computed results from the screening runs. The performance metrics in the table below were calculated using the `screen_analyzer()` function. The key metrics are:

*   **Recall**: The proportion of truly relevant studies that the model correctly identified. High recall is crucial to avoid missing relevant studies.
*   **Specificity**: The proportion of truly irrelevant studies that the model correctly identified.
*   **Balanced Accuracy (bAcc)**: The average of recall and specificity, providing a single measure that balances performance on both relevant and irrelevant studies.

```{r}
#| echo: false
#| eval: true
table_path <- system.file("extdata", "comparison-results", "comparing_gpt41.csv", package = "AIscreenR")

results_table <- read.csv(table_path, check.names = FALSE)
knitr::kable(results_table, align = "lcccc")
```

## Interpretation

The results show clear differences in model behavior depending on both architecture and the number of repetitions. With a single repetition, gpt-4.1-mini achieves the highest balanced accuracy (0.958) and recall (0.938), while gpt-4.1 offers the highest specificity (0.983) but lower recall (0.844) and balanced accuracy (0.914). gpt-4.1-nano performs less well in recall and balanced accuracy when run once.

Increasing the number of repetitions to ten for the mini and nano models leads to modest changes. For gpt-4.1-mini recall decreases slightly (0.922), and balanced accuracy drops to 0.949. For gpt-4.1-nano recall improves (0.828 vs. 0.719), and balanced accuracy increases to 0.902.

The main focus should be on maximizing recall to avoid missing relevant studies. gpt-4.1-mini with a single repetition provides the best overall recall and balanced accuracy. gpt-4.1-nano benefits from more repetitions, improving recall and balanced accuracy, but still lags behind mini and full models. The full gpt-4.1 model, while having the highest specificity, shows lower recall and balanced accuracy, making it less suitable for screening tasks where recall is critical.

It should be noted that the decrease in recall for gpt-4.1-mini when increasing repetitions might be due to the small number of relevant studies compared to irrelevant ones, meaning that even a single misclassification can significantly impact recall. Thus, while more repetitions can help, they do not always guarantee better recall. Therefore, it is recommended to prioritize a precise prompt with clear instructions to get the best performance especially for bigger models like gpt-4.1-mini or gpt-4.1.
