---
title: "Comparing GPT-5 Models for Screening"
format:
  html:
    code-fold: false
    code-tools: true
    code-summary: "Show the code"
    toc: true
execute: 
  echo: fenced
  warning: false
  message: false
knitr:
  opts_chunk: 
    fig.pos: "H"
    fig.retina: 2
    cache: FALSE
    R.options:
      knitr.graphics.auto_pdf: true
      width: 100
      knitr.kable.NA: "-"
      dplyr.summarise.inform: FALSE
      scipen: 10
      pillar.sigfig: 4 
editor_options: 
  chunk_output_type: console
---

## Introduction
To understand the difference between the different GPT-5 models and the effect of `reasoning_effort`, we compare the performance of `gpt-5`, `gpt-5-mini`, and `gpt-5-nano` on a real-world dataset. We will analyze their performance with a single repetition (`reps = 1`). For the faster and more economical `gpt-5-mini` and `gpt-5-nano` models, we will also investigate the impact of increasing the screening to ten repetitions (`reps = 10`). We will be using a section of the `friends_dat` consisting of 100 studies, 80 randomly sampled studies marked as 'exclude' and 20 randomly sampled studies marked as 'include'.

## Setup
First, we load the necessary packages for our analysis.
```{r}
#| echo: true
#| eval: false

library(AIscreenR)
library(dplyr)
library(knitr)
```
## Methodology: Running the Comparison

To compare the models, we performed a screening task on the section of the `friends_dat` dataset. The goal was to identify studies about the FRIENDS preventive programme.

The following prompt was used for all models. It is designed to be very specific about the required output format to ensure consistent results.

```{r}
#| echo: true
#| eval: false

# This is the prompt used for the experiment.
prompt <- "You are a systematic literature review screening agent. You are equipped with a screening function tool that you MUST use.

MANDATORY BEHAVIOR:
- Call the provided screening function tool immediately after reading each study
- Do not provide any response other than the function call
- The function call is your complete and only response

SCREENING CONTEXT:
Topic: FRIENDS preventive programme's effect on reducing anxiety symptoms in children/adolescents
Criteria 1: Study about FRIENDS preventive programme. The FRIENDS programme is a 10-session manualised cognitive behavioural therapy (CBT) programme which can be used as both prevention and treatment of child and youth anxiety. 
Criteria 2: Study is estimating an effect between a treatment and control/comparison group

WORKFLOW:
1. Read title and abstract
2. Make function call with your assessments
3. End response immediately after function call

CRITICAL: Your entire response must be the function tool call. No additional text before, during, or after the function call."
```

For each model and `reasoning_effort` combination, we run the screening using the `tabscreen_gpt()` function from the `AIscreenR` package. We set the `model` parameter to the respective GPT-5 model and the `screening_effort` parameter to `low`, `medium`, or `high` then we vary the `reps` parameter to 1 or 10.

```{r}
#| echo: true
#| eval: false

# This code was run for each model and repetition setting.
# It is not evaluated here to save time, as we will load the results directly.
plan(multisession)

results <- 
  tabscreen_gpt(
    data = friends_dat, 
    prompt = prompt, 
    model = "gpt-5", # This was changed for each model
    reps = 1, # This was changed to 10 for the second run
    reasoning_effort = "low", # This was changed to "medium" and "high" for other runs
  )

plan(sequential)

# The results were saved to an .rds file for later analysis.
saveRDS(results, "result_obj_gpt-5_rep1.rds")
```

## Results

We now load the pre-computed results from the screening runs. The performance metrics in the table below were calculated using the `screen_analyzer()` function. The key metrics are:

*   **Recall**: The proportion of truly relevant studies that the model correctly identified. High recall is crucial to avoid missing relevant studies.
*   **Specificity**: The proportion of truly irrelevant studies that the model correctly identified.
*   **Balanced Accuracy (bAcc)**: The average of recall and specificity, providing a single measure that balances performance on both relevant and irrelevant studies.
```{r}
#| echo: false
#| eval: true

table_path <- system.file("extdata", "comparison-results", "comparing_gpt5.csv", package = "AIscreenR")

if (!nzchar(table_path) || !file.exists(table_path)) {
  stop("Could not find comparing_gpt5.csv in the installed package extdata.")
}

results_table <- read.csv(table_path, check.names = FALSE)
knitr::kable(results_table, align = "lcccc")
```

## Interpretation
The results show notable differences in model behavior depending on both architecture, `reasoning_effort`, and the number of repetitions. With a single repetition and high effort, gpt-5-mini achieves the highest balanced accuracy (0.844) and recall (0.75), correctly identifying 15 of 20 relevant studies. In contrast, gpt-5 and gpt-5-nano show more modest performance, with their best balanced accuracy reaching 0.800 and recall of 0.65.

Increasing the `reasoning_effort` has a meaningful impact for gpt-5-mini, where recall improves from 0.65 (low) to 0.75 (high). For gpt-5, moving from low to medium effort improves recall (0.60 to 0.65), but no further gains are seen at high effort. For gpt-5-nano, `reasoning_effort` has minimal impact, with recall remaining largely flat at 0.60 across effort levels.

Ten repetitions had mixed effects. For gpt-5-mini, adding repetitions left performance unchanged at low effort but reduced recall at medium  and high effort, lowering the balanced accuracy for both. For gpt-5-nano, repetitions had no impact at low effort and yielded small gains at medium and high effort: recall increased from 0.60 to 0.65, raising bAcc from 0.787 to 0.800 (medium) and 0.775 to 0.800 (high), with a specificity trade-off at medium. Overall, increasing repetitions does not consistently improve screening quality and can reduce recall in some settings, particularly for gpt-5-mini. Increasing repetitions seems more beneficial for smaller noisier models like gpt-5-nano.

The main focus for systematic review screening should be on maximizing recall to avoid missing relevant studies. gpt-5-mini with high effort and a single repetition provides the best overall recall and balanced accuracy. gpt-5-nano shows limited responsiveness to both effort and repetitions, making it less suitable when recall is critical. The gpt-5 model does not outperform gpt-5-mini, suggesting that a larger model may not be necessary for this screening task.

With only 20 relevant studies, even a single additional false negative significantly impacts recall (each study represents 5 percentage points). Thus, while more repetitions can theoretically reduce noise, they do not guarantee better recall and may in fact reduce it. Therefore, it is recommended to prioritize higher `reasoning_effort` with a single repetition, combined with a precise prompt with clear instructions, to achieve the best performance.