---
title: "Reading and Writing RIS Files"
format:
  html:
    code-fold: false
    code-tools: true
    code-summary: "Show the code"
    toc: true
vignette: |
  %\VignetteIndexEntry{Reading and Writing RIS Files}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
execute: 
  echo: fenced
  warning: false
  message: false
knitr:
  opts_chunk: 
    fig.pos: "H"
    fig.retina: 2
    cache: FALSE
    R.options:
      knitr.graphics.auto_pdf: true
      width: 100
      knitr.kable.NA: "-"
      dplyr.summarise.inform: FALSE
      scipen: 10
      pillar.sigfig: 4 
editor_options: 
  chunk_output_type: console
---

::: {.callout-important}
These functions were developed because existing R tools for loading RIS files often fail to preserve the original formatting from PsycINFO. The new functions are designed to handle this RIS file structure with other database formats. For an full-scale use of these functions, see the [Using OpenAI’s GPT API models for Title and Abstract Screening in Systematic Reviews](Using-GPT-API-Models-For-Screening.html) vignette. If you experience any issues with these functions, please report them on [GitHub](https://github.com/MikkelVembye/AIscreenR/issues). Otherwise, you can try using [`synthesisr::read_refs()`](https://martinwestgate.com/synthesisr/reference/read_refs.html).
:::

## Overview

This vignette introduces two helpers for working with RIS files. The function `read_ris_to_dataframe(file_path)` parses an RIS file into a data frame with the following features:

- Automatically maps RIS tags to descriptive column names (e.g., `AU` → `author`, `TI` → `title`, `PY` → `year`)
- Preserves the order of tags as they first appear in the file
- Collapses repeated tags within a record into a single semicolon-separated string
- Stores metadata to preserve original formatting when writing back

The function `save_dataframe_to_ris(df, file_path)` writes a data frame back to RIS format:

- Writes `TY` (source type) first for each record, followed by all other fields
- Splits semicolon-separated values into multiple RIS tag lines
- Preserves original formatting when available (from metadata)
- Terminates each record with `ER  - `


## Load the package

```{r}
#| echo: true
#| eval: true

library(AIscreenR)
```

## Read an RIS file

The example below builds a small RIS file in a temporary location and reads it.

```{r}
#| echo: true
#| eval: true

ris <- c(
  "TY  - JOUR",
  "AU  - Author, One",
  "AU  - Author, Two",
  "TI  - An example title",
  "PY  - 2020",
  "ER  - ",
  "",
  "TY  - CHAP",
  "TI  - Another title",
  "AU  - Author, Three",
  "ER  - "
)

tmp_in <- tempfile(fileext = ".ris")
writeLines(ris, tmp_in, useBytes = TRUE)

df <- read_ris_to_dataframe(tmp_in)
df
```

The output data frame has descriptive column names instead of RIS tags. For example:

- `TY` (type) becomes `source_type`
- `AU` (author) becomes `author`
- `TI` (title) becomes `title`
- `PY` (publication year) becomes `year`

Repeated tags, such as multiple `AU` lines, are collapsed to a single string with "; " (e.g., "Author, One; Author, Two").

## Write a data frame to RIS

Create a data frame and write it to a `.ris` file. You can use either descriptive column names (as returned by `read_ris_to_dataframe()`) or raw RIS tags. Semicolon-separated values are automatically split into multiple tag lines.

```{r}
#| echo: true
#| eval: true

# Using raw RIS tags
df_out <- data.frame(
  TY = c("JOUR", "CHAP"),
  AU = c("Author, One; Author, Two", "Author, Three"),
  TI = c("An example title", "Another title"),
  PY = c("2020", ""),
  stringsAsFactors = FALSE
)

tmp_out <- tempfile(fileext = ".ris")
invisible(capture.output(save_dataframe_to_ris(df_out, tmp_out)))
readLines(tmp_out, encoding = "UTF-8")
```

Each record writes the `TY` tag first, splits any field value containing ";" into multiple RIS tag lines, and ends with `ER  - ` followed by a blank line.

You can also use descriptive column names (they will be automatically mapped back to RIS tags):

```{r}
#| echo: true
#| eval: true

# Using descriptive names
df_descriptive <- data.frame(
  source_type = c("JOUR", "CHAP"),
  author = c("Author, One; Author, Two", "Author, Three"),
  title = c("An example title", "Another title"),
  year = c("2020", ""),
  stringsAsFactors = FALSE
)

tmp_out2 <- tempfile(fileext = ".ris")
invisible(capture.output(save_dataframe_to_ris(df_descriptive, tmp_out2)))
readLines(tmp_out2, encoding = "UTF-8")
```

Both approaches produce identical RIS output.
