---
title: "Reading and Writing RIS Files"
format: html
vignette: >
  %\VignetteIndexEntry{Reading and Writing RIS Files}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

## Overview

This vignette introduces two helpers for working with RIS files. The function `read_ris_to_dataframe(file_path)` parses an RIS file into a data frame with the following features:

- Automatically maps RIS tags to descriptive column names (e.g., `AU` → `author`, `TI` → `title`, `PY` → `year`)
- Preserves the order of tags as they first appear in the file
- Collapses repeated tags within a record into a single semicolon-separated string
- Stores metadata to preserve original formatting when writing back

The function `save_dataframe_to_ris(df, file_path)` writes a data frame back to RIS format:

- Writes `TY` (source type) first for each record, followed by all other fields
- Splits semicolon-separated values into multiple RIS tag lines
- Preserves original formatting when available (from metadata)
- Terminates each record with `ER  - `

```{r}
library(AIscreenR)
```

## Read an RIS file

The example below builds a small RIS file in a temporary location and reads it.

```{r}
ris <- c(
  "TY  - JOUR",
  "AU  - Author, One",
  "AU  - Author, Two",
  "TI  - An example title",
  "PY  - 2020",
  "ER  - ",
  "",
  "TY  - CHAP",
  "TI  - Another title",
  "AU  - Author, Three",
  "ER  - "
)

tmp_in <- tempfile(fileext = ".ris")
writeLines(ris, tmp_in, useBytes = TRUE)

df <- read_ris_to_dataframe(tmp_in)
df
```

The output data frame has descriptive column names instead of RIS tags. For example:

- `TY` (type) becomes `source_type`
- `AU` (author) becomes `author`
- `TI` (title) becomes `title`
- `PY` (publication year) becomes `year`

Repeated tags, such as multiple `AU` lines, are collapsed to a single string with "; " (e.g., "Author, One; Author, Two").

## Write a data frame to RIS

Create a data frame and write it to a `.ris` file. You can use either descriptive column names (as returned by `read_ris_to_dataframe()`) or raw RIS tags. Semicolon-separated values are automatically split into multiple tag lines.

```{r}
# Using raw RIS tags
df_out <- data.frame(
  TY = c("JOUR", "CHAP"),
  AU = c("Author, One; Author, Two", "Author, Three"),
  TI = c("An example title", "Another title"),
  PY = c("2020", ""),
  stringsAsFactors = FALSE
)

tmp_out <- tempfile(fileext = ".ris")
invisible(capture.output(save_dataframe_to_ris(df_out, tmp_out)))
readLines(tmp_out, encoding = "UTF-8")
```

Each record writes the `TY` tag first, splits any field value containing ";" into multiple RIS tag lines, and ends with `ER  - ` followed by a blank line.

You can also use descriptive column names (they will be automatically mapped back to RIS tags):

```{r}
# Using descriptive names
df_descriptive <- data.frame(
  source_type = c("JOUR", "CHAP"),
  author = c("Author, One; Author, Two", "Author, Three"),
  title = c("An example title", "Another title"),
  year = c("2020", ""),
  stringsAsFactors = FALSE
)

tmp_out2 <- tempfile(fileext = ".ris")
invisible(capture.output(save_dataframe_to_ris(df_descriptive, tmp_out2)))
readLines(tmp_out2, encoding = "UTF-8")
```

Both approaches produce identical RIS output.
